% En esta sección introducimos las nociones formales que utilizaremos a lo largo del trabajo. Primero definimos los \textit{Shapley values} y su instanciación al problema de \textit{feature attribution}. Luego presentamos los \textit{Asymmetric Shapley Values} (ASV), que incorporan relaciones entre las variables a la atribución mediante un grafo causal. Finalmente, introducimos las redes bayesianas como modelo para la distribución de los datos y describimos las familias restringidas de redes que consideraremos en nuestros resultados.

%\subsection{Shapley values y feature attribution}

Let $X$ be a finite set of features. A (binary) \textit{entity} over $X$ is a mapping $e:X \to \{0, 1\}$ assigning a value to every feature from $X$. We denote by $\entities{}(X)$ the set of all entities, and we will assume that some probability distribution is provided over this set.
A (binary) classifier is a mapping $M: \entities{}(X) \to \{0, 1\}$. We restrict to both binary entities and classifiers only for simplicity: all the results and algorithms we obtain can be extended to more general cases straightforwardly (unless we say otherwise). 

Given a model $M$ and an entity $e$, a \textit{feature attribution score} is a mapping $\phi_{M,e}:X \to \R$. Intuitively, the value $\phi_{M,e}(x)$ indicates the relevance of feature $x$ with respect to prediction $M(e)$. One of the most 
studied feature attribution schemes is the \SHAPscore{} \parencite{shapOriginalPaper}, which is theoretically grounded on the \emph{Shapley values} \parencite{shapley1953value} from cooperative game theory. In that context, the Shapley values
come up as the ``most fair'' way to distribute some capital earned during a team game. More precisely, given a finite set of players $\players$ and a \emph{characteristic function} $\charactheristicFunction:\mathcal{P}(\players)\to\mathbb{R}$ describing, for each subset
of the players, the reward that such coalition would obtain in some fixed game; the shapley values are the unique assignment $\varphi_\nu: \players \to \R$ of reward to each player $i$ satisfying the following desired properties:

\begin{enumerate}
	\item Efficiency: All the reward is distributed, i.e. $\sum_{i \in \players} \varphi_{\nu}(i) = \nu(\players)$.
	\item Symmetry: If $\nu(S \cup \{i\}) = \nu(S \cup \{j\})$ for every $S \subseteq \players$, then $\varphi_{\nu}(i) = \varphi_{\nu}(j)$.
	\item Null Player: If $\nu(S \cup \{i\}) = 0$ for every $S \subseteq \players$, then $\varphi_{\nu}(i) = 0$.
	\item Linearity: For any pair of characteristic functions $\nu_1,\nu_2$ and $a \in \R$ it holds that $\varphi_{a\nu_1 + \nu_2}(i) = a\varphi_{\nu_1}(i) + \varphi_{\nu_2}(i)$\footnote{The characteristic function $a \nu_1 + \nu_2$ is naturally defined as $(a \nu_1 + \nu_2)(S) = a \nu_1(S) + \nu_2(S)$.}. 
\end{enumerate}

Moreover, the Shapley values have an exact closed form: if we refer to players with indices in the range $\{1,\ldots, n\}$, we can write
%
\begin{align}\label{eq:shapley}
	\varphi_{\nu}(i) =  \frac{1}{|\players|!} \sum_{\pi \in perm(\players)} \left[ \nu(\pi_{<i} \cup \{i\}) - \nu(\pi_{<i}) \right]
\end{align}
%
where $perm(\players)$ denotes the set of all permutations of the number from 1 to $n$ and $\pi_{<i}$ denotes the set of indices before $i$ in the permutation $\pi$ (i.e. $\pi_{<i} = \{j : \pi(j) < i\}$). Fundamentally, Eq.~\eqref{eq:shapley} is summing across all possible ways
of ordering the features, considering the impact of the inclusion of player $i$ to the reward of the game when adding the players according to each of these orderings.

In the context of machine learning explainability, the Shapley values are instantiated by thinking of the features as players and considering the characteristic function
%
\begin{align}\label{eq:char_fun_for_ML}
	\nu_{M,e}(S) = \mathbb{E}[M(e') | S, e] = \sum_{e'\in \entities(X)} M(e') \Pr[e'| \{e'(x) = e(x) : x \in S\}] 
\end{align}
%
which represents the expected value of the model $M$ when restricting the features from $S$ to have the value dictated by entity $e$.\footnote{In some proof we may write the characteristic function as $\nu_{M,e,\Pr}$ to make the dependence on the distribution $\Pr$ explicit.} 
Note that the properties that these values satisfy do not have an evident satisfactory meaning in the context of explainability \parencite{fryer2021shapley}. Furthermore, reasonable requisites for \textit{feature attribution} methods related to the theory of \textit{abductive explanations} are not entailed by these properties \parencite{huang2023inadequacy, marques2023logic}. Nonetheless, the Shapley values and the related SHAP-score are useful tools for explaining model behaviour in real scenarios \parencite{lundberg2020local}.

Depending on the family of ML models under consideration and the underlying distribution of the entities the computation of the Shapley values can be tractable or not. For example,
it is well-known that for product distributions\footnote{A product distribution over $\entities{}(X)$ is a distribution that factorizes as $\Pr\left[e\right] = \prod_{x \in X}\Pr\left[ x = e(x) \right]$.} the complexity of computing these scores is equivalent to the complexity of computing the average 
value of the model \parencite{van2022tractability}. In particular, this implies that there exists efficient algorithms for the case where the models are given, for example, by decision trees or by restricted families of circuits \parencite{arenas2023complexity}, while the problem is intractable whenever the models are more expressive, as in the case of neural nets. 
For more general families of distributions the problem becomes much harder: for instance, when considering Naive Bayes distributions, the computation of this score is intractable already when considering trivial models that only depend on one feature \parencite{van2022tractability}. Even though there are particular distributions that generalize the product one for which efficient algorithms are known \parencite{marzouk2025computationaltractabilitymanyshapley}
the general applicability of these tools seems to depend on heuristic approximations of the scores.

Another problem of the Shapley values is that they may fail to distinguish features that are heavily correlated. For example, if some feature $j$ depends deterministically of some other feature $i$ it can happen that both features receive the same score, even though $i$ is intuitivelly more relevant
for \textit{any} model (PONER EJEMPLO?). Note that the \textit{Symmetry} property essentially implies that this should always happen whenever two features share the same information relevant for a prediction. Thus, to avoid this problem it is somewhat necessary to relax that axiom and therefore define a different score.
This is the idea underlying the \textit{Assymetric Shapley Values} (ASV from now on).

\begin{definition}[Assymetric Shapley Values \parencite{frye2019asymmetric}]\label{def:asv}
	Given a model $M$, an entity $e$, a feature $x$ and some weight function $w: perm(X) \to \R_{\geq 0}$ satisfying 
	\begin{align*}
		\sum_{\pi \in perm(X)} w(\pi) = 1,
	\end{align*}
	%
	we define the Assymetric Shapley value of $x$ as
	%
	\begin{align*}
		\phi^{ASV}_{M,e}(x) = \sum_{\pi \in perm(X)} w(\pi) \left[ \nu_{M,e}(\pi_{<i} \cup \{i\}) - \nu_{M,e}(\pi_{<i}) \right]
	\end{align*}
	%
	where $\nu_{M,e}$ is defined in Eq.~\eqref{eq:char_fun_for_ML} 
\end{definition}

This definition is derived from the corresponding notion of assymetric shapley values from cooperative game theory. The function $w$ is useful for weighting some orderings above others. For example, if some feature $j$ is a function of $i$ it is possible to introduce this fact into the score by assigning weight zero to all orderings in which $i$
comes after $j$. Note that the Shapley values are obtained by taking $w(\pi) = \frac{1}{|X|!}$, and we will refer to this particular case as $\phi^{Shap}_{M,e}$.

\cite{frye2019asymmetric} proposes to define the weight function $w$ by considering a \textit{causal DAG} $C$ over the features. More precisely, $C$ is a directed acyclic graph whose nodes are the set of features $X$ and where an edge directed from $i$ to $j$ indicates that $j$ depends on $i$. Then, they define the weight function
%
\begin{align}\label{eq:w_in_ASV}
	w_C(\pi) = \begin{cases}
		\frac{1}{|topos(C)|} & \pi \text{ is a topological sort of } C\\
		0 & \text{otherwise}
	\end{cases}	
\end{align}
%
This weight function generalizes the idea mentioned earlier of considering only those orderings consistent with the known causality of the features. In the rest of this paper we will be interested in developing efficient algorithms for computing the ASV from Def.~\ref{def:asv} whenever the weight function is given by 
Eq.~\eqref{eq:w_in_ASV}. 

Moreover, we will consider that the distribution of space of entites is described by Bayesian Network $N$. We refer to the reader to standard bibliography for the precise formal definitions of Bayesian Networks \parencite{pearl1986bayesianInference} while here we describe them conceptually: a bayesian network $N$
defines a distribution for $\entities(X)$ with a DAG whose nodes are the features from $X$. Every node $x$ has a set of parents $p(x) \subseteq X$, and the network provides, for each way of assigning the features in $p(x)$, the conditional distribution of $x$. See (EJEMPLO) for an example. Given $N$ and any entity $e$ it is possible to compute
the probability of $e$ by traversing the DAG of $N$ in any topological sort multiplying the corresponding conditional probabailities (EL EJEMPLO PODRIA MOSTRAR ESTO UN TOQUE?).

One of the most simples types of Bayesian Networks is the Naive Bayes, which was already mentioned in regards to intractibility results for the Shapley values. This network has a particular structure, in which a single node is parent to all the others and no other pairs of nodes are connected. See () for an example.

Sometimes we will identity the network itself as a causal graph. Thus, the network will provide both the distrubition necessary to compute the characteristic function $\nu_{M,e}$ as well as the causal graph necessary to define the weight function $w^C$. (AGREGAR DISCUSION DE CUANDO ESTO ESTA BIEN Y CUANDO NO)

(Capaz podemos poner el cacho comentado abajo donde se describen formalmente las redes bayesianas. Pero veamos si podemos hacernos entender con solo esto.)

\begin{comment}
\subsection{Asymmetric Shapley Values y grafos causales}

\echu{No hablamos mucho de la complejidad de Shap o de ASV, lo queremos hacer acá y contar los resultados previos o la idea es no contarlo y decirlo de una en los resultados teóricos?}

La definición anterior trata a todas las permutaciones de features de manera uniforme. Sin embargo, si se dispone de un grafo causal sobre las variables, resulta natural restringir la atención a aquellas permutaciones compatibles con dicha estructura.

La generalización más simple consiste en reemplazar el promedio uniforme por una función de pesos arbitraria sobre las permutaciones.

\begin{definition}[Shapley values ponderados]
	Sea $w:\perm(\players)\to\mathbb{R}$ una función tal que $\sum_{\pi\in\perm(\players)} w(\pi)=1$. Definimos
	\begin{align}
		\phi_i^{assym}(\charactheristicFunction)
		=
		\sum_{\pi\in\perm(\players)}
		w(\pi)
		\left[
		\charactheristicFunction(\pi_{<i}\cup\{i\})-
		\charactheristicFunction(\pi_{<i})
		\right].
		\label{eq:assymetric_shap_def_definitions}
	\end{align}
\end{definition}

La familia de \emph{Asymmetric Shapley Values} propuesta en \cite{frye2019asymmetric} surge al elegir la función de pesos a partir de un DAG que codifica relaciones causales entre las features.

\begin{definition}[Asymmetric Shapley Values]
	Sea $G=(X,E)$ un DAG cuyos nodos son los features. Denotemos por $\topo(G)$ al conjunto de órdenes topológicos de $G$. Definimos la función de pesos
	\[
	w(\pi)=
	\begin{cases}
		\frac{1}{|\topo(G)|} & \text{si }\pi\in\topo(G),\\
		0 & \text{en otro caso}.
	\end{cases}
	\]
	Entonces, dado un modelo $M$, una entidad $e$ y una distribución $\Pr$, el valor ASV de un feature $x_i\in X$ se define como
	\[
	\assym_{M,e,\Pr}(x_i)
	=
	\frac{1}{|\topo(G)|}
	\sum_{\pi\in\topo(G)}
	\left[
	\charFunML_{M,e,\Pr}(\pi_{<i}\cup\{x_i\})-
	\charFunML_{M,e,\Pr}(\pi_{<i})
	\right].
	\]
\end{definition}

La intuición detrás de esta definición es que sólo se consideran órdenes que respetan la causalidad. De este modo, si una variable $x$ es causa de una variable $y$, entonces $x$ siempre aparece antes que $y$ en las permutaciones admitidas. Esto tiende a favorecer a las variables causalmente anteriores y a reducir la importancia de variables que son meramente consecuencias de otras.

\subsection{Redes bayesianas como modelo de distribución}

Para evaluar los ASV no alcanza con fijar un modelo predictivo: también es necesario especificar una distribución sobre el espacio de entidades. En este trabajo utilizamos redes bayesianas para representar dicha distribución.

\begin{definition}[Red bayesiana]
	Una Red Bayesiana $\aBayesianNetwork$ sobre un conjunto de variables $X$ es una tupla
	$\aBayesianNetwork=(X,E,\Pr)$, donde $(X,E)$ es un DAG y, para cada variable $x\in X$, se especifica una distribución condicional $\Pr(x\mid \parents(x)).$
	Estas distribuciones se representan mediante tablas de probabilidad condicional (\emph{Conditional Probability Tables}, CPTs).
\end{definition}

La propiedad central de una red bayesiana es que factoriza la distribución conjunta usando únicamente dependencias locales. Si $X=\{X_1,\dots,X_n\}$, entonces
\[
\Pr(X_1,\dots,X_n)=\prod_{i=1}^n \Pr(X_i\mid \parents(X_i)).
\]
Esta factorización permite realizar inferencia probabilística de forma mucho más eficiente que manipulando directamente la distribución conjunta.

Conviene distinguir con cuidado entre \emph{red causal} y \emph{red bayesiana}. Una red bayesiana modela dependencias condicionales y define una distribución conjunta; un grafo causal, en cambio, pretende reflejar relaciones de causa y efecto entre variables. En general, ambas nociones no coinciden necesariamente. Sin embargo, en este trabajo asumiremos que el DAG subyacente a la red bayesiana puede utilizarse también como grafo causal para caclular los ASV. Esta decisión responde a dos motivos: por un lado, es consistente con el tipo de datos y experimentos que analizamos; por otro, nos permite estudiar simultáneamente el costo de la inferencia probabilística y el efecto de imponer restricciones causales en el cálculo de explicaciones.

Más precisamente, dado un modelo $M$, una entidad $e$ y una red bayesiana $\aBayesianNetwork$ sobre las features, utilizaremos la distribución inducida por $\aBayesianNetwork$ para calcular la función característica $\charFunML_{M,e,\Pr}$, y el DAG de $\aBayesianNetwork$ como estructura causal para definir $\assym_{M,e,\Pr}$.

\subsection{Familias restringidas de redes bayesianas}

La complejidad de la inferencia en redes bayesianas depende fuertemente de la estructura del grafo subyacente. En redes generales, calcular probabilidades marginales o condicionadas es un problema $\sharpPhard$, mientras que para ciertas familias restringidas la inferencia es tratable. Por esta razón nos centramos en grafos para los cuáles la inferencia puede realizarse en tiempo polinomial.

La primera familia que consideramos es la de las redes \emph{Naive Bayes}. En este caso, existe una variable raíz $X_1$ tal que todas las demás variables son sus hijas, y no hay otras aristas. Es decir, el conjunto de aristas es $ E=\{(X_1,X_j):2\leq j\leq n\}.$.
La distribución conjunta factoriza entonces como
\[
\Pr[e]=\Pr[X_1=e(X_1)]\prod_{j=2}^n \Pr[X_j=e(X_j)\mid X_1=e(X_1)].
\]

\echu{Hace falta mostrar la fórmula de la probabilidad o mejor mostrar la imagen de que estructure tiene y listo? Igual la imagen ocupa espacio y no suma tanto, tal vez no hace falta ninguna de las dos.}
Esta familia es especialmente relevante porque, aunque su estructura es muy simple, ya permite exhibir una diferencia fundamental entre SHAP y ASV: se sabe que calcular SHAP sobre distribuciones Naive Bayes es intratable incluso para modelos muy simples \cite{van2022tractability}, mientras que uno de nuestros resultados muestra que ASV puede ser tratable en este escenario.

La segunda familia central en este trabajo es la de los \emph{polytrees}, es decir, DAGs cuyo grafo subyacente no dirigido es un árbol. Esta clase resulta importante porque admite inferencia probabilística en tiempo polinomial \cite{pearl1986bayesianInference}. Como el cálculo de la función característica involucra predicciones promedio condicionadas a evidencia, la posibilidad de realizar inferencia eficiente en la distribución de entrada es un ingrediente fundamental para obtener algoritmos tratables para ASV.

\echu{¿Se dice tratable o tractable?}

En síntesis, a lo largo de esta tesis trabajaremos principalmente con dos tipos de restricciones estructurales sobre la distribución de los datos: redes Naive Bayes, que sirven como caso base para nuestros resultados de tractabilidad, y polytrees, que constituyen la clase más general sobre la que desarrollamos algoritmos eficientes para inferencia y cómputo de explicaciones.

\echu{¿Este parrafo juega o este de más?}
\end{comment}