In this work, we studied \textit{Asymmetric Shapley Values} (ASV), a variant of Shapley values that incorporates causal information into feature attribution. We analyzed the computational complexity of ASV and showed tractability results for restricted settings, including Naive Bayes and bounded-degree polytrees. To make ASV more practical, we developed exact and approximate algorithms based on equivalence classes of topological orderings, as well as efficient procedures for mean prediction in decision trees under Bayesian network distributions. Our experiments show that these ideas significantly reduce the number of required evaluations and yield accurate approximations in tractable cases.

Overall, our results suggest that ASV is a promising explainability framework when causal structure is relevant, since it can capture dependencies that standard SHAP does not reflect. At the same time, the work highlights that the main challenge is computational rather than conceptual: ASV can be made practical in structured settings, but further optimization is needed to make it competitive at larger scale.

Several directions remain open for future work. On the theoretical side, it would be interesting to generalize the equivalence class algorithm from directed trees to polytrees, and to further study the complexity of counting and sampling topological orderings. On the practical side, possible improvements include implementing alternative counting algorithms from the literature, optimizing the current implementation through caching or parallelization, and extending the framework to more general causal models beyond Bayesian networks.