\documentclass{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%\urlstyle{rm}
%
\usepackage[spanish, es-noshorthands]{babel}
\addto\captionsspanish{\renewcommand{\tablename}{Tabla}} %Para que diga Tabla en vez de Cuadro
\usepackage{dsfont, amsmath}

\usepackage{caption}
\usepackage{subcaption}  % Para tener captions dentro de las subfiguras
\captionsetup{font=small}%para que las captions de las figuras sean más chicas

\usepackage{amssymb}
\usepackage{booktabs} % For better table formatting
\usepackage{graphicx} % Required for inserting images
\usepackage[colorlinks=true, urlcolor=blue, linkcolor=blue, citecolor=green]{hyperref}
%linkColor = citas internas cómo ref
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.pathmorphing}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{pgf} % For arithmetic operations in TikZ
\usepackage{comment}
\usepackage{booktabs}
\usepackage{url} %For writing paths
\usepackage[normalem]{ulem}
\usepackage{multirow}
\newcommand{\scc}[1]{{\color{red}#1}}


\usepackage[style=apa]{biblatex} % <- AGREGAR ESTA LÍNEA
\addbibresource{biblio.bib} % <- AGREGAR ESTA LÍNEA

\input{secciones/macros}

\tikzset{
	nodo/.style={
		shape=circle,
		draw=black,
		line width=1pt,   
		minimum size=7mm
	},
	arista/.style={
		line width=1pt,  
		-{Latex[length=3mm]}  
	},  
	mySnake/.style={
		decorate, 
		decoration={snake, amplitude=.4mm, segment length=4mm, post length=1mm}
	}
}


\begin{document}
%
% ENGLISH VERSION
\title{Contribution title}

%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Princeton University, Princeton NJ 08544, USA \and
Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{lncs@springer.com}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words. This is the English version of the abstract that 
provides a comprehensive overview of the research conducted, 
the methodology used, and the main findings obtained.

% keywords in lowercase
\keywords{first keyword\and second keyword\and another keyword}
\end{abstract}

% SPANISH VERSION
% Clear the previous title info and set Spanish version

% Spanish title and abstract
\begin{center}
\vspace{1cm}
{\Large \bfseries\boldmath Título de la contribución}

\vspace{0.5cm}
\end{center}

\begin{abstract}
\noindent En este artículo se aborda el problema de la explicabilidad en modelos de aprendizaje automático mediante métodos de feature attribution.

En particular, se estudia una variante de los Shapley values conocida como Asymmetric Shapley Values (ASV), que permite incorporar conocimiento causal en la explicación de modelos de forma model-agnostic. A partir del análisis de su complejidad, se demuestra que el cálculo exacto de ASV es polinomial en modelos cuya distribución de entrada está representada por una red bayesiana del tipo Naive Bayes, en contraste con SHAP, que es \#P-hard aún en este caso restringido.

Con el objetivo de extender estos resultados a clases más generales de redes bayesianas, se introduce una noción de clases de equivalencia sobre los órdenes topológicos del grafo causal subyacente, lo cual permite reducir drásticamente el número de permutaciones necesarias para computar ASV. Se presenta un algoritmo polinomial en el número de clases para identificarlas, y se implementa un esquema de cómputo exacto de ASV basado en estas clases. Además, se propone un nuevo método para computar en tiempo polinomial la predicción esperada de un árbol de decisión, sobre una distribución dada por una red bayesiana arbitraria, permitiendo así evaluar el algoritmo desarrollado para el cómputo de ASV en estos modelos. 

Por último, se propone un algoritmo aproximado para calcular el ASV en familias de DAG's causales del tipo polytree. Para ello, se desarrolla un algoritmo de muestreo aleatorio de órdenes topológicos de polytrees. Estos resultados respaldan la viabilidad del enfoque propuesto en estructuras causales realistas, y se contrastan empíricamente con SHAP tanto en precisión como en eficiencia computacional.
\bigskip


\keywords{ explicabilidad, árboles de decisión, Asymmetric Shapley Values (ASV), Shap, polytrees, ordenes topológicos}
\end{abstract}


\vspace{0.5cm}
\setcounter{page}{1}

%
\section*{TODO}

\begin{itemize}
    \item Escribir todo xD
\end{itemize}

\section{Introducción}

\input{secciones/1_introduccion}

\section{Definiciones}

\input{secciones/2_definiciones}

\section{Theoretical results}

\begin{enumerate}
    \item Resultado de que ASV se puede calcular en algunos casos donde el otro no se podía (el caso se la naiva bayes digo). Creo que no hay apendice, asi que diria de, en principio, poner todas 
    las demos en el main text. Tienen que estar prolijas.
    \item Idea de las clases de equivalencia. Definición y relación con la cantidad de órdenes topológicos.
    \item Descripción general del algoritmo basado en clases de equivalencias para cómputo exacto.
    \item Instanciar este algoritmo en el caso de árboles y dar la complejidad.
    \item Introducir la opción de samplear para aproximar el ASV, tanto clases de equivalencia como ordenes topológicos.
    \item Dar dos resultados generales diciendo "Si podes samplear clases / ordenes en X tiempo y calcular promedios en Y, la complejidad es f(XY)".
    \item Instanciar este resultado en casos particulares. Hablar de los algoritmos para promedios de arboles. Capaz el resultado de imposibilidad.
    \item Dar el resultado basado en sampleo de ordenes topologicos en general.
\end{enumerate}

\section{Experimental results}

\begin{enumerate}
    \item Comparación entre clases de equivalencia y ordenes topologicos.
    \item Algun resultado de performance sobre algoritmos de clases de equivalencias. 
    \item Si podemos, estaria bueno implementar el otro algoritmo basado en sampleo de ordenes topologicos.
    \item No me parece necesario contar la performance de los algoritmos que calculan promedios, aunque si hay lugar podemos.
    \item Si hay lugar tambien podriamos hacer una comparacion de shapley y ASV (no me queda claro si hay o no pq son 14 paginas pero el formato es una verga)
\end{enumerate}

\section{Conclusions}

\section{First Section}
\subsection{A Subsection Sample}
Please note that the first paragraph of a section or subsection is
not indented. The first paragraph that follows a table, figure,
equation etc. does not need an indent, either.

Subsequent paragraphs, however, are indented.

\subsubsection{Sample Heading (Third Level)} Only two levels of
headings should be numbered. Lower level headings remain unnumbered;
they are formatted as run-in headings.

\paragraph{Sample Heading (Fourth Level)}
The contribution should contain no more than four levels of
headings. Table~\ref{tab1} gives a summary of all heading levels.

\begin{table}
\caption{Table captions should be placed above the
tables.}\label{tab1}
\begin{tabular}{|l|l|l|}
\hline
Heading level &  Example & Font size and style\\
\hline
Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
\hline
\end{tabular}
\end{table}


\noindent Displayed equations are centered and set on a separate
line.
\begin{equation}
x + y = z
\end{equation}
Please try to avoid rasterized images for line-art diagrams and
schemas. Whenever possible, use vector graphics instead (see
Fig.~\ref{fig1}).

\begin{theorem}
This is a sample theorem. The run-in heading is set in bold, while
the following text appears in italics. Definitions, lemmas,
propositions, and corollaries are styled the same way.
\end{theorem}
%
% the environments 'definition', 'lemma', 'proposition', 'corollary',
% 'remark', and 'example' are defined in the LLNCS documentclass as well.
%
\begin{proof}
Proofs, examples, and remarks have the initial word in italics,
while the following text appears in normal font.
\end{proof}
For citations of references, we use the APA 7 convention.  The following bibliography provides
a sample reference list with entries for journal
articles~\parencite{ref_article1}, an LNCS chapter~\parencite{ref_lncs1}, a
book~\parencite{ref_book1}, proceedings without editors~\parencite{ref_proc1},
and a homepage~\parencite{ref_url1}. Multiple citations are grouped
\parencite{ref_article1,ref_lncs1,ref_book1},
\parencite{ref_article1,ref_book1,ref_proc1,ref_url1}.

\begin{credits}
\subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
used for general acknowledgments, for example: This study was funded
by X (grant number Y).

\subsubsection{\discintname}
It is now necessary to declare any competing interests or to specifically
state that the authors have no competing interests. Please place the
statement with a bold run-in heading in small font size beneath the
(optional) acknowledgments,
for example: The authors have no competing interests to declare that are
relevant to the content of this article. Or: Author A has received research
grants from Company W. Author B has received a speaker honorarium from
Company X and owns stock in Company Y. Author C is a member of committee Z.
\end{credits}
%
% ---- Bibliography ----
%
\printbibliography

\end{document}